Addition and removal energies of circular quantum dots

### 2017-07-01 Manuscript/paper.tex
Forgot factorial in Slater det
Tweak and simplify style

\input{head_revtex.tex}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{color}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{scalerel}
\usepackage{siunitx}

@@

\author{Fei Yuan}
\email{yuan@nscl.msu.edu}
\homepage{https://people.nscl.msu.edu/~yuan}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Samuel J. Novario}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Nathan Parzuchowski}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{The Ohio State University, Columbus, OH 43210, USA}
\author{Sarah Reimann}
\affiliation{Department of Chemistry and Center for Theoretical and Computational Chemistry, University of Oslo, N-0316 Oslo, Norway}
\author{Scott K. Bogner}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Morten Hjorth-Jensen}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}

@@

The behavior of strongly confined electrons is a major problem for many-body theory.  The prototypical system is that of quantum dots, also known as ``artificial atoms'', in which electrons are confined within artificially constructed semiconducting heterostructures.  Such nanoscale systems are highly relevant as they lie at a critical point where both theoretical models and experimental observations can readily probe quantum phenomena such as tunneling, entanglement, and magnetization \cite{reimann2002,engel1993}.  Moreover, unlike many physical systems, quantum dots enjoy the benefit of being highly tunable through changes in the external field or the structure of the confining material.  This allows experiments to quantify the impact of quantum effects at different levels of correlation.

@@

In this paper, we analyze the application of several methods to the quantum dot systems, including the Hartree--Fock (HF) method, M\o ller-Plesset perturbation theory (MP), the in-medium similarity renormalization group (IM-SRG) method, coupled cluster (CC) theory \cite{PhysRevB.67.045320,heidari:114708,PhysRevB.84.115302}, quasidegenerate perturbation theory (QDPT), and equations-of-motion (EOM) methods.  We compare them against variational and diffusion Monte Carlo (VMC and DMC) \cite{PhysRevB.68.035304,PhysRevB.62.8120,PhysRevB.84.115302,PhysRevB.54.4780} and full configuration interaction (FCI) theory \cite{olsen2013thesis,JJAP.36.3924,PhysRevB.56.6428,2008arXiv0810.2644K,rontani:124102} results, where available, from the existing literature.

@@

\begin{figure}
  \includegraphics{figures/fig-shell-structure-v2.pdf}
  \caption{(Color online) The 42 lowest single-particle states (the first 5 shells) in the 2D harmonic oscillator basis.  Each box represents a single-particle state arranged by $m_\ell$, $m_{\mathrm{s}}$, and energy, and the up/down arrows indicate the spin of the states.  Within each column, the principal quantum number $n$ increases as one traverses upward.}
  \label{fig:shell-structure}
\end{figure}

The noninteracting part of the many-body Hamiltonian $\hat{H}_1$ can be reduced to $N$ independent single-particle Hamiltonians each of the form:
\begin{align*}
  \hat{h} = \frac{1}{2} \hat{\bm{\nabla}}^2 + \frac{1}{2} \omega^2 \hat{\bm{r}}^2.
\end{align*}
Therefore, solving the noninteracting many-body problem amounts to solving the single-particle time-independent Schr\"odinger equation
\begin{align*}
  \hat{h} |p\rangle = \varepsilon_p |p\rangle,
\end{align*}
where $p$ is an unknown tuple of quantum numbers.  In Cartesian coordinates, the two-dimensional harmonic oscillator is trivially reducible to the well-known one-dimensional problem.  However, to exploit the circular symmetry, we prefer to use Fock-Darwin states, which are written in polar coordinates $\bm{r} = (\rho, \varphi)$.  Such states have the computationally useful property of conserving orbital angular momentum $\hat{L}_{\mathrm{z}} \equiv -\mathrm{i} \frac{\partial}{\partial \varphi}$.

The Fock-Darwin wave functions can be decomposed into radial and angular components,\cite{lohne2010coupled}
\begin{align}
  F_{n m_\ell}(\rho, \varphi) &\equiv \sqrt\omega R_{n |m_\ell|}(\sqrt \omega \rho) A_{m_\ell}(\varphi) \label{eq:fockdarwin}, \\
  R_{n m}(\varrho) &\equiv \sqrt{\frac{2 \cdot n!}{(n + m)!}} \mathrm{e}^{-\varrho^2 / 2} \varrho^m L_n^{(m)}(\varrho^2), \notag \\
  A_{m_\ell}(\varphi) &\equiv \frac{1}{\sqrt{2 \pi}} \mathrm{e}^{\mathrm{i} m_\ell \varphi}, \notag
\end{align}
where $L_n^{(\alpha)}$ denotes the (generalized) Laguerre polynomial \cite{NIST:DLMF} of degree $n$ and parameter $\alpha$,
\begin{align*}
  L_n^{(\alpha)}(u) = \frac{1}{n!} u^{-\alpha} \mathrm{e}^u \frac{\mathrm{d}^n}{\mathrm{d} u^n} (\mathrm{e}^{-u} u^{\alpha + n}).
\end{align*}
The states are labeled by two quantum numbers: the principal quantum number $n \in \{0, 1, 2, \ldots\}$ and orbital angular momentum projection $m_\ell \in \mathbb \{\ldots, -2, -1, 0, +1, +2, \ldots\}$.  For a wave function, $n$ indicates the degree of the Laguerre polynomial, whereas $m_\ell$ is simply the eigenvalue of $\hat{L}_{\mathrm{z}}$.

@@

Returning to the one-body part of the many-body Hamiltonian, $\hat{H}_1$, we can use the Slater determinant construction to build an $N$-particle eigenstate $|p_1, \ldots, p_N\rangle$ of the many-body Hamiltonian $\hat{H}_1$ from an arbitrary choice of $N$ single-particle eigenstates $|p_1\rangle, \ldots, |p_N\rangle$ from \eqref{eq:singleparticlestate}, where $p$ is an abbreviation for the triplet $(n, m_\ell, m_{\mathrm{s}})$.  Note that a Slater determinant $|p_1 \ldots p_N\rangle$ is not merely a product of the single-particle states, but an antisymmetrized sum over all possible products,
\begin{align*}
  \langle \bm{r}_1 \ldots \bm{r}_N | p_1 \ldots p_N \rangle \equiv
  \frac{1}{\sqrt{N}} \left|
  \begin{matrix}
    \langle \bm{r}_1 | p_1 \rangle & \cdots & \langle \bm{r}_1 | p_N \rangle \\
    \vdots & \ddots & \vdots \\
    \langle \bm{r}_N | p_1 \rangle & \cdots & \langle \bm{r}_N | p_N \rangle \\
  \end{matrix}
  \right|.
\end{align*}
The Slater determinant $|p_1 \ldots p_N\rangle$ is said to \textit{occupy} the states $|p_1\rangle, \ldots, |p_N\rangle$.  By construction, Slater determinants satisfy the Pauli exclusion principle, ensuring that the exchange of any two particles is equivalent to a reversal in the sign of the many-body wave function.

@@

\begin{figure}
  \centering
  \includegraphics{figures/fig-methods.pdf}
  \caption{(Color online) A schematic view of the various ways in which many-body methods in this paper could be combined to calculate ground state, addition, and removal energies.}
  \label{fig:methods}
\end{figure}

@@

\begin{table}
  \centering
  \caption{Ground state energy of quantum dots with $N$ particles and an oscillator frequency of $\omega$.  For every row, the calculations are performed in a harmonic oscillator basis with $K$ shells.  The abbreviation ``n.c.'' stands for ``no convergence'': these are cases where IM-SRG(2) or CCSD either diverged or converged extremely slowly.}
  \label{tab:ground}
  \input{tab-ground}
\end{table}

@@

\begin{table}
  \centering
  \caption{Addition energy of quantum dot systems.  See Table \ref{tab:ground} for details.}
  \label{tab:add}
  \input{tab-add}
\end{table}

\begin{table}
  \centering
  \caption{Removal energy of quantum dot systems.  See Table \ref{tab:add} for details.}
  \label{tab:rm}
  \input{tab-rm}
\end{table}

@@

\begin{table}
  \centering
  \caption{Extrapolated ground state energies for quantum dots with fit uncertainties, computed from the approximate Hessian in the Levenberg--Marquardt fitting algorithm.  These uncertainties also determine the number of significant figures presented.  Extrapolations are done using 5-point fits where the number of shells $K$ ranges between $K_{\text{stop}} - 4$ and $K_{\text{stop}}$ (inclusive).  The abbreviation ``n.c.'' stands for ``no convergence'': these are extrapolations where, out of the 5 points, at least one of them was unavailable because IM-SRG(2) or CCSD either diverged or converged extremely slowly.}
  \label{tab:ground-extrapolated}
  \input{tab-ground-extrapolated}
\end{table}

\begin{table}
  \centering
  \caption{Extrapolated addition energies for quantum dots with fit uncertainties.  The abbreviation ``n.c.'' has the same meaning as in Table \ref{tab:ground-extrapolated}.  The abbreviation``n.f.'' stands for ``no fit'': this particular extrapolation resulted in unphysical parameters ($\beta \le 0$).  See Table \ref{tab:ground-extrapolated} for other details.}
  \label{tab:add-extrapolated}
  \input{tab-add-extrapolated}
\end{table}

\begin{table}
  \centering
  \caption{Extrapolated removal energies for quantum dots with fit uncertainties.  See Table \ref{tab:add-extrapolated} for details.}
  \label{tab:rm-extrapolated}
  \input{tab-rm-extrapolated}
\end{table}

@@

We note that this calculation was done entirely using the traditional approach of using a high-order ODE solver to solve the flow equation.  A new technique developed by T. D. Morris \cite{PhysRevC.92.034331} uses an alternative approach that obviates the need for a high-order ODE solver, leading to much more efficient computations and also allowing operators of other observables to be evolved at lower cost, which poses a significant advantage over CC methods.  Implementing this approach would allows us to study the accuracy and convergence of other possibly more sensitive observables.

@@

Implementation-wise, our current code constructs the two-particle states as simple Slater determinants of their one-particle states.  This approach is often referred to as \textit{m-scheme} in nuclear physics.  It is straightforward to implement and works well for general systems, but it fails to exploit all the symmetries in the quantum dot system: not only is spin projection $\hat S_z$ conserved, the Casimir operator $\hat S^2$ is as well.  A more efficient approach is to couple the spins of the single-particle states to form two-particle states with good $\hat S^2$, an approach analogous to \textit{j-scheme} in nuclear physics.  A combination of this with the Eckart-Wigner theorem could reduce the computational cost significantly, albeit at the cost of increased implementation complexity.

@@

\bibliographystyle{\mybibstyle}
\bibliography{paper}
\end{document}

>>>>

\input{head_revtex.tex}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{scalerel}
\usepackage{siunitx}

@@

\author{Fei Yuan}
\email{yuan@nscl.msu.edu}
\homepage{https://people.nscl.msu.edu/~yuan}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Samuel J. Novario}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Nathan M. Parzuchowski}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{The Ohio State University, Columbus, OH 43210, USA}
\author{Sarah Reimann}
\affiliation{Department of Chemistry and Center for Theoretical and Computational Chemistry, University of Oslo, N-0316 Oslo, Norway}
\author{S. K. Bogner}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Morten Hjorth-Jensen}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}

@@

The behavior of strongly confined electrons is a major problem for many-body theory.  The prototypical system is that of quantum dots, also known as ``artificial atoms'', in which electrons are confined within artificially constructed semiconducting heterostructures.  Such nanoscale systems are highly relevant as both theoretical models and experimental observations can readily probe quantum phenomena such as tunneling, entanglement, and magnetization \cite{reimann2002,engel1993}.  Moreover, unlike many physical systems, quantum dots enjoy the benefit of being highly tunable through changes in the external field or the structure of the confining material.  This allows experiments to quantify the impact of quantum effects at different levels of correlation.

@@

In this paper, we analyze the application of several methods to the quantum dot systems, including the Hartree--Fock (HF) method \cite{hartree_1928,Fock1930}, M\o ller--Plesset perturbation theory (MP) \cite{1934PhRv...46..618M}, the in-medium similarity renormalization group (IM-SRG) method \cite{IMSRG}, coupled cluster (CC) theory \cite{PhysRevB.67.045320,heidari:114708,PhysRevB.84.115302}, quasidegenerate perturbation theory (QDPT) \cite{0022-3700-7-18-010,Kvasnicka1974}, and equations-of-motion (EOM) methods \cite{RevModPhys.40.153}.  We compare them against variational and diffusion Monte Carlo (VMC and DMC) \cite{PhysRevB.68.035304,PhysRevB.62.8120,PhysRevB.84.115302,PhysRevB.54.4780} and full configuration interaction (FCI) theory \cite{olsen2013thesis,JJAP.36.3924,PhysRevB.56.6428,2008arXiv0810.2644K,rontani:124102} results, where available, from the existing literature.

@@

\begin{figure}
  \includegraphics{fig-shell-structure-v2.pdf}
  \caption{(Color online) The 42 lowest single-particle states (the first 5 shells) in the 2D harmonic oscillator basis.  Each box represents a single-particle state arranged by $m_\ell$, $m_{\mathrm{s}}$, and energy, and the up/down arrows indicate the spin of the states.  Within each column, the principal quantum number $n$ increases as one traverses upward.}
  \label{fig:shell-structure}
\end{figure}

The noninteracting part of the many-body Hamiltonian $\hat{H}_1$ can be reduced to $N$ independent single-particle Hamiltonians each of the form:
\begin{align*}
  \hat{h} = \frac{1}{2} \hat{\bm{\nabla}}^2 + \frac{1}{2} \omega^2 \hat{\bm{r}}^2.
\end{align*}
Therefore, solving the noninteracting many-body problem amounts to solving the single-particle time-independent Schr\"odinger equation
\begin{align*}
  \hat{h} |p\rangle = \varepsilon_p |p\rangle,
\end{align*}
where $p$ is an unknown tuple of quantum numbers.  In Cartesian coordinates, the two-dimensional harmonic oscillator is trivially reducible to the well-known one-dimensional problem.  However, to exploit the circular symmetry, we prefer to use Fock--Darwin states, which are written in polar coordinates $\bm{r} = (\rho, \varphi)$.  Such states have the computationally useful property of conserving orbital angular momentum $\hat{L}_{\mathrm{z}} \equiv -\mathrm{i} \frac{\partial}{\partial \varphi}$.

The Fock--Darwin wave functions can be decomposed into radial and angular components,\cite{lohne2010coupled}
\begin{align}
  F_{n m_\ell}(\rho, \varphi) &\equiv \sqrt\omega R_{n |m_\ell|}(\sqrt \omega \rho) A_{m_\ell}(\varphi) \label{eq:fockdarwin}, \\
  R_{n m}(\varrho) &\equiv \sqrt{\frac{2 \times n!}{(n + m)!}} \mathrm{e}^{-\varrho^2 / 2} \varrho^m L_n^{(m)}(\varrho^2), \notag \\
  A_{m_\ell}(\varphi) &\equiv \frac{1}{\sqrt{2 \pi}} \mathrm{e}^{\mathrm{i} m_\ell \varphi}, \notag
\end{align}
where $L_n^{(\alpha)}$ denotes the (generalized) Laguerre polynomial \cite{NIST:DLMF} of degree $n$ and parameter $\alpha$,
\begin{align*}
  L_n^{(\alpha)}(u) = \frac{1}{n!} u^{-\alpha} \mathrm{e}^u \frac{\mathrm{d}^n}{\mathrm{d} u^n} (\mathrm{e}^{-u} u^{\alpha + n}).
\end{align*}
The states are labeled by two quantum numbers: the principal quantum number $n \in \{0, 1, 2, \ldots\}$ and orbital angular momentum projection $m_\ell \in \mathbb \{\ldots, -2, -1, 0, +1, +2, \ldots\}$.  For a wave function, $n$ indicates the degree of the Laguerre polynomial, whereas $m_\ell$ is simply the eigenvalue of $\hat{L}_{\mathrm{z}}$.

@@

Returning to the one-body part of the many-body Hamiltonian, $\hat{H}_1$, we can use the Slater determinant construction to build an $N$-particle eigenstate $|p_1, \ldots, p_N\rangle$ of the many-body Hamiltonian $\hat{H}_1$ from an arbitrary choice of $N$ single-particle eigenstates $|p_1\rangle, \ldots, |p_N\rangle$ from \eqref{eq:singleparticlestate}, where $p$ is an abbreviation for the triplet $(n, m_\ell, m_{\mathrm{s}})$.  Note that a Slater determinant $|p_1 \ldots p_N\rangle$ is not merely a product of the single-particle states, but an antisymmetrized sum over all possible products,
\begin{align*}
  \langle \bm{r}_1 \ldots \bm{r}_N | p_1 \ldots p_N \rangle \equiv
  \frac{1}{\sqrt{N!}} \left|
  \begin{matrix}
    \langle \bm{r}_1 | p_1 \rangle & \cdots & \langle \bm{r}_1 | p_N \rangle \\
    \vdots & \ddots & \vdots \\
    \langle \bm{r}_N | p_1 \rangle & \cdots & \langle \bm{r}_N | p_N \rangle \\
  \end{matrix}
  \right|.
\end{align*}
The Slater determinant $|p_1 \ldots p_N\rangle$ is said to \textit{occupy} the states $|p_1\rangle, \ldots, |p_N\rangle$.  By construction, Slater determinants satisfy the Pauli exclusion principle, ensuring that the exchange of any two particles is equivalent to a reversal in the sign of the many-body wave function.

@@

\begin{figure}
  \centering
  \includegraphics{fig-methods.pdf}
  \caption{(Color online) A schematic view of the various ways in which many-body methods in this paper could be combined to calculate ground state, addition, and removal energies.}
  \label{fig:methods}
\end{figure}

@@

\begin{table}
  \centering
  \caption{Ground state energy of quantum dots with $N$ particles and an oscillator frequency of $\omega$.  For every row, the calculations are performed in a harmonic oscillator basis with $K$ shells.  The abbreviation ``n.c.'' stands for ``no convergence'': these are cases where IM-SRG(2) or CCSD either diverged or converged extremely slowly.}
  \label{tab:ground}
  \input{tab-ground}\unskip
\end{table}

@@

\begin{table}
  \centering
  \caption{Addition energy of quantum dot systems.  See Table \ref{tab:ground} for details.}
  \label{tab:add}
  \input{tab-add}\unskip
\end{table}

\begin{table}
  \centering
  \caption{Removal energy of quantum dot systems.  See Table \ref{tab:add} for details.}
  \label{tab:rm}
  \input{tab-rm}\unskip
\end{table}

@@

\begin{table}
  \centering
  \caption{Extrapolated ground state energies for quantum dots with fit uncertainties, computed from the approximate Hessian in the Levenberg--Marquardt fitting algorithm.  These uncertainties also determine the number of significant figures presented.  Extrapolations are done using 5-point fits where the number of shells $K$ ranges between $K_{\text{stop}} - 4$ and $K_{\text{stop}}$ (inclusive).  The abbreviation ``n.c.'' stands for ``no convergence'': these are extrapolations where, out of the 5 points, at least one of them was unavailable because IM-SRG(2) or CCSD either diverged or converged extremely slowly.}
  \label{tab:ground-extrapolated}
  \input{tab-ground-extrapolated}\unskip
\end{table}

\begin{table}
  \centering
  \caption{Extrapolated addition energies for quantum dots with fit uncertainties.  The abbreviation ``n.c.'' has the same meaning as in Table \ref{tab:ground-extrapolated}.  The abbreviation``n.f.'' stands for ``no fit'': this particular extrapolation resulted in unphysical parameters ($\beta \le 0$).  See Table \ref{tab:ground-extrapolated} for other details.}
  \label{tab:add-extrapolated}
  \input{tab-add-extrapolated}\unskip
\end{table}

\begin{table}
  \centering
  \caption{Extrapolated removal energies for quantum dots with fit uncertainties.  See Table \ref{tab:add-extrapolated} for details.}
  \label{tab:rm-extrapolated}
  \input{tab-rm-extrapolated}\unskip
\end{table}

@@

We note that this calculation was done entirely using the traditional approach of using a high-order ODE solver to solve the flow equation.  A new technique developed by Morris et al.\ \cite{PhysRevC.92.034331} uses an alternative approach based on the Magnus expansion that obviates the need for a high-order ODE solver, leading to much more efficient computations and also allowing operators of other observables to be evolved at lower cost, which poses a significant advantage over CC methods.  Implementing this approach would allows us to study the accuracy and convergence of other possibly more sensitive observables.

@@

Implementation-wise, our current code constructs the two-particle states as simple Slater determinants of their one-particle states.  This approach is often referred to as \textit{m-scheme} in nuclear physics.  It is straightforward to implement and works well for general systems, but it fails to exploit all the symmetries in the quantum dot system: not only is spin projection $\hat S_z$ conserved, the Casimir operator $\hat S^2$ is as well.  A more efficient approach is to couple the spins of the single-particle states to form two-particle states with good $\hat S^2$, an approach analogous to \textit{j-scheme} in nuclear physics.  A combination of this with the Wigner--Eckart theorem could reduce the computational cost significantly, albeit at the cost of increased implementation complexity.

@@

\bibliography{paper}
\end{document}
